---
title: "Introduction to hydroroute"
author: "Julia Haider"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to hydroroute}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = 'center'
)
old <- options(useFancyQuotes = FALSE)
library("hydroroute")
```

## Introduction

The method developed by Greimel et al. (2016) detects and
characterizes sub-daily flow fluctuations and is implemented in the R
package hydropeak (available on
[CRAN](https://CRAN.R-project.org/package=hydropeak)). Based on the
events detected by the method implemented in package hydropeak,
hydroroute identifies associated events in hydrographs from
neighboring gauging stations and models the translation and retention
processes between them (Greimel et al., 2022).

This vignette presents the main function `peaktrace()` which given the
events and relation information between gauging stations determines
the associated events and based on these, estimates predictive models
to trace initial values specified for the relevant metrics across the
neighboring gauging stations. First, an overview on the input data
required is given and the additional function `get_lag()` and
variants thereof are presented which estimate the mean translation time of
the hydrographs between adjacent gauging stations, if not already
available. Then the individual function `estimate_AE()` is presented
which is used by `peaktrace()` to identify the associated events
(AEs). Finally the application of `peaktrace()` is illustrated and it
is shown how the return value can be inspected.

## Input data

Several data files are required to perform the analysis. 

### Dataset `Q`

If the mean translation time between hydrographs at neighboring
gauging stations is not given or known, the raw dataset `Q`,
containing (equispaced) date-time values and the corresponding flow
fluctuations, is needed. Based on these data the mean translation time
between hydrographs at neighboring gauging stations may be estimated
using `get_lag()`. The dataset `Q` needs to contain three variables:

1. `ID` Character string which refers to the identifier of the gauging station 
    (in Austria: HZBCODE).
2. `Time` Character string with date-time information of stage measurements.
3. `Q` Numeric, flow (stage measurements).

The combination of `ID` and `Time` must be unique. Functions that use
the dataset `Q` assume that these variables are contained in this
order with exactly these names.  If this is not the case, i.e., if the
columns have different names or are not in this order, the order can
be specified in these functions with argument `cols`.  The columns are
then renamed internally to make the data processable. Also the
date-time format must be specified if it is different from the
function's default format used.

The following code loads the sample dataset `Q` and shows the first few rows:
```{r}
Q_path <- system.file("testdata", "Q.csv", package = "hydroroute")
Q <- read.csv(Q_path)
head(Q)
```

The sample dataset `Q` is a data frame with `r nrow(Q)` rows and
`r ncol(Q)` variables as described above. The station ID's along the flow
path are ``r paste(sort(unique(Q$ID)), collapse = ", ")``.
The time ranges from ``r dQuote(min(Q$Time))`` to ``r dQuote(max(Q$Time))``.

### Dataset `relation`

The dataset `relation` provides information about the gauging stations
of neighboring hydrographs. It contains:

1. `ID` Character string which refers to the identifier of the gauging station 
    (in Austria: HZBCODE).
2. `Type` Character string which characterizes the source hydrograph
    (`Turbine flow`, `Gauge`, `Basin outflow`).
3. `Station` Character string which indicates the order of the `n` hydrographs 
    in `relation` in downstream direction (`Si` with `i = 1, ..., n`).
4. `fkm` Numeric, position of hydrograph in km relative to the source.
5. `LAG` Character string which contains the cumulative mean
    translation time (or estimated cumulative lag) between the source
    and a specific gauging station in the format `HH:MM`. For `S1`
    this is either indicated as missing (`NA`) or always given as
    `00:00`. It is either already provided in `relation` or can be
    estimated from the corresponding dataset `Q` with `get_lag()`.

The following code loads an example dataset:

```{r}
relation_path <- system.file("testdata", "relation.csv", package = "hydroroute")
relation <- read.csv(relation_path)
relation
```

The dataset `relation` contains `r nrow(relation)` adjacent gauging stations.

### Event files

The output files from hydropeak's `get_events_*()` function are used
to identify AEs. The naming scheme of the output files is
`ID_event-type_date-time-from_date-time-to.csv`.  Event types are
defined as follows:

- 0: Constant event after an NA event (i.e., a missing event) or as first event in the time series.
- 1: Constant event after a decreasing event.
- 2: Increasing event (IC).
- 3: Constant event after an increasing event.
- 4: Decreasing event (DC).
- 5: NA event.

The most important event types for the following analysis are `2`
(increasing event; IC) and `4` (decreasing event; DC).

Package hydroroute includes 8 sample `Event` files for each gauging
station ID contained in the sample dataset `Q` and event type `2` (IC)
and `4` (DC) between `"2014-01-01 00:00:00"` and `"2014-02-28
23:45:00"`. The increasing events for the station with ID `100000` are
thus loaded using:

```{r}
Sx <- system.file("testdata", "Events", "100000_2_2014-01-01_2014-02-28.csv",
                  package = "hydroroute")
Sx <- read.csv(Sx)
head(Sx)
```

## Get mean translation time between hydrographs with `get_lag()`

For the identification of AEs, the translation time between
neighboring hydrographs and the event amplitude have to be
considered. For the first criterion, the mean translation time (`LAG`)
between hydrographs has to be estimated and the cumulative values
appended to the `relation` data for further processing, if not
available yet.

Function `get_lag_file()` uses:

- `Q_file` A path to a file that contains the `Q` data from several
  stations or a data frame that contains this information.
- `relation_file` A path to a `relation` file. The `ID`s of the stations
  must be in `Q`.

If the argument `save` is `TRUE`, the `relation` data with appended
`LAG` column is written to a file specified in `outfile`. If a `LAG`
column already exists, argument `overwrite` has to be set to `TRUE` to
overwrite the existing column. The function can be applied to several
`relation` files by iterating over file paths or if a single `Q` data
file is available, `get_lag_dir()` can be used. `relation` files can
be selected from a directory using regular expressions (argument
`relation_pattern`).

The following code shows this for single file names:
```{r}
Q_file <- system.file("testdata", "Q.csv", package = "hydroroute")
relation <- system.file("testdata", "relation.csv", package = "hydroroute")
get_lag_file(Q_file, relation, inputsep = ",", format = "%Y-%m-%d %H:%M",
             save = FALSE, overwrite = TRUE)
```

This code indicates the use with `get_lag_dir()` where the directory
is specified:

```{r}
Q_file <- system.file("testdata", "Q.csv", package = "hydroroute")
relations_path <- system.file("testdata", package = "hydroroute")
get_lag_dir(Q_file, relations_path, inputsep = ",",
            inputdec = ".", format = "%Y-%m-%d %H:%M", overwrite = TRUE)
```

## Estimate settings with `estimate_AE()`

Greimel et al. (2022) propose the following algorithm to identify AEs:

"Starting from every single event at the upstream hydrograph (`Sx`), the
mean translation time between the hydrographs [...] is subtracted from
the downstream hydrograph (`Sy`). This then captures several events from
the downstream hydrograph within a time slot similar to the
translation time. The event detected at `Sy` with the amplitude most
similar to `Sx` is designated as potential AE. [...]

For these events, then the relative difference in amplitude is
determined. Parabolas are fitted to the histogram obtained for the relative 
difference data binned into intervals from -1 to 1 of width 0.1 by fixing the 
vertex at the inner maximum of the histogram and the width is determined by
minimizing the average squared distance between the parabola and the
histogram data along arbitrary symmetric ranges from the inner
maximum. Based on the fitted parabola cut points with the x-axis are
determined such that only those potential AEs are retained where the
relative difference is within these cut points. If this automatic
scheme does not succeed to determine suitable cut points, e.g.,
because the estimated cut points are outside -1 and 1, then a strict
criterion for the relative difference in amplitude is imposed to
identify AEs considering only deviations of at most 10%."

`estimate_AE()` estimates suitable settings for the amplitude based on
the method developed in Greimel et al. (2022) for exact time matches.

It performs this procedure for two neighboring hydrographs, i.e., it
takes a subset of `relation` and the two corresponding `Event` files
as input.  The gauging station `ID`s in the subset of `relation` and
in the `Event` files must match.  Suitable settings for the amplitude
are estimated as follows:

- `Sy$Time` is shifted by the mean translation time between `Sx` and
  `Sy`. 
  
- Based on the time, exact matches between `Sx` and `Sy`
  are captured.

- Relative difference in `AMP` are computed, e.g., `(Sy$AMP -
  Sx$AMP) / Sx$AMP`.
  
- Matched events detected at `Sy` with the amplitude most similar to
  `Sx` are filtered. However, only matched events with an amplitude at
  `Sy` of at most two times the amplitude at `Sx` are considered as
  potential AEs.
    
- The relative differences of potential AEs are binned into intervals of 
  length 0.1 from -1 to 1. The created relative frequency table of the binned
  relative differences is passed to function `get_parabola()` where
  either suitable cut points with the x-axis are determined or a
  strict criterion is returned.
    
- The table of the relative differences is visualized in a plot where
  the fitted parabola and the cut points with the x-axis are also
  shown.
  
- The estimated settings for amplitude, a data frame of "real" AEs,
  i.e., associated events within the estimated cut points, and the
  plot are returned.
  
Note that the metric flow ratio (RATIO) does not make sense for `S1`
if the hydrograph is not of type `Gauge`. So metric `RATIO` is set to
`NA` internally in this case.
    
The following code shows this procedure for two `Event` files:
```{r}
# file paths
Sx <- system.file("testdata", "Events", "100000_2_2014-01-01_2014-02-28.csv",
                  package = "hydroroute")
Sy <- system.file("testdata", "Events", "200000_2_2014-01-01_2014-02-28.csv",
                  package = "hydroroute")
relation <- system.file("testdata", "relation.csv", package = "hydroroute")

# read data
Sx <- utils::read.csv(Sx)
Sy <- utils::read.csv(Sy)
relation <- utils::read.csv(relation)
relation <- relation[1:2, ]

# estimate AE, exact time matches
results <- estimate_AE(Sx, Sy, relation, timeLag = c(0, 1, 0))
```

```{r}
results$settings
```
Column `bound` represents the lower, inner and upper bounds that are used to subset
potential AEs. `lag` represents the time lag. Only exact matches are used in this
examples, which is specified by argument `timeLag = c(0, 1, 0)`, which refers to 
0 deviation at the lower and upper bound and 1 at the inner bound, meaning, that 
the mean translation time from `relation` is not altered when time matches are 
computed. 


```{r, fig.width = 6, fig.height = 3}
results$plot_threshold
```

```{r}
head(results$real_AE)
```

## Combine everything with `peaktrace()`

Function `peaktrace` combines the identification of potential AEs and
the estimation of suitable amplitude settings for a whole river
section as specified in a `relation` file. In addition, the flow
metrics of the AEs are pictured by scatter plots and the translation
and retention process between the hydrographs is described by linear
models.

In the following, the input arguments of function `peaktrace()` are
described:

- `relation_path`: Is the path where the whole relation file from a
  river section is to be read from.
    
- `events_path`: Is the path of the directory where the `Event` files
  are located. These files must correspond to the format described
  earlier in the section discussing the input data.
    
- `initial_values_path`: Is the path where initial values for
  predicting the metric at the neighboring stations are to be read
  from. It should not contain missing values. But missing values can be imputed
  with a method specified in argument `impute_method`, which is by default `max`.
  An example for such a file is:


```{r}
initial_values_path <- system.file("testdata", "initial_value_routing.csv",
                                   package = "hydroroute")
initials <- read.csv(initial_values_path)
initials
```

The columns must be identical to this example. The content may
vary. The initial values used for prediction must not contain any
missing values. 

- `Station` refers to the gauging station of the hydrograph. Here, all
  initial values correspond to gauging station `S1` except for metric
  `RATIO`, which starts at gauging station `S2`.
    
- `Metric` corresponds to the metric. This is used to pick the
  corresponding fitted predictive model.
    
- `Value` can be chosen arbitrarily or estimated with a data-driven
  approach. A unique `Name` is assigned which can be used to
  characterize the curve obtained from this initial value used to
  predict the metrics in downstream direction. E.g., for this example
  the initial values are set to certain quantiles of the metrics at
  station `S1`.

The return value of function `peaktrace()` is structured as follows:

- A named list with one list element for each `event_type`.
- For each `event_type`:

  1. One element that contains the estimated settings from
     `estimate_AE()` for all gauging stations.
     
  2. Plot of relative differences of `AMP` with cut points from
     settings_AE()` for all pairs of neighboring gauging stations.
     
  3. Real AEs according to the estimated settings from `estimate_AE()`
     for all pairs of neighboring gauging stations and a column
     `diff_metric` that contains the relative difference in `AMP`.
    
  4. A grid of scatter plots containing the AEs for neighboring
     hydrographs and for each metric with the fitted regression line.
	
  5. Results of model fitting. Each row contains the corresponding
     stations and metric, the model type (default: "lm"), formula,
     coefficients, number of observations and $R^2$.
	
  6. Plot of predicted values based on the initial values. 

```{r}
relation_path <- system.file("testdata", "relation.csv", package = "hydroroute")
events_path <- system.file("testdata", "Events", package = "hydroroute")
initial_values_path <- system.file("testdata", "initial_value_routing.csv",
                                   package = "hydroroute")
res <- peaktrace(relation_path, events_path, initial_values_path)
```

The first list object refers to event type 2 (IC event). 

```{r}
res$`2`$settings
```

The `settings` data frame contains the estimated time lag and metric
settings computed with `estimate_AE()`. In this example with 4
stations, it contains nine rows where three rows describe the relation
between two neighboring stations. Since only exact time matches were
allowed, the `lag` values are 0 for `lower` and `upper` bound and 1 for `inner`. 
`metric` contains the range of relative values of the amplitude allowed.  
Events, where the relative difference in amplitude and the relative difference 
in time are within these settings, are considered as "real" AE and therefore events
caused by disruptive factors are excluded as far as possible.


The following plot shows the histograms obtained for the relative
differences in amplitude for each pair of neighboring gauging stations
binned into intervals from -1 to 1 of width 0.1. The dashed line shows
the fitted parabola and the cut points of the parabola with the x-axis
are indicated. Potential AEs where the relative difference is within
these cut points are considered as "real" AEs.

```{r, fig.width = 8, fig.height = 3}
grid::grid.draw(res$`2`$plot_threshold)
```

The "real" AEs can be insepcted using:

```{r}
head(res$`2`$real_AE)
```

The scatter plots of the metrics at the neighboring gauging stations
for the "real" AEs are contained in `plot_scatter`. The scatter plots
are arranged in a grid where each row contains scatter plots for a
specific metric and each colums contains a different pair of
neighboring gauging stations. The x-axis is the upstream hydrograph
`Sx`, the y-axis is the downstream hydrograph `Sy`. A linear
regression line and the corresponding $R^2$ value are added to each
plot. By default the aspect ratio is fixed and the axis limits are
equal within each plot.

```{r, fig.width = 8, fig.height = 15}
grid::grid.draw(res$`2`$plot_scatter)
```

The fitted regression models may also be inspected:

```{r}
res$`2`$models
```

The `models` data frame contains the fitted (linear) models for each pair of 
neighboring stations and each metric. 

- `station.x` is the upstream hydrograph `Sx`.
- `station.y` is the downstream hydrograph `Sy`.
- `metric` is the name of the corresponding metric.
- `type` is the model class, by default `lm`.
- `formula` is the expression that is used to fit the model.
- `(Intercept)`, `x` are the extracted coefficients (called with `coef`).
- `n` is the number of events used to fit the model.
- `r2` is the extracted or computed $R^2$ for each model.

Finally the fitted regression models are used to predict the values of
the metrics along the longitudinal flow path given the initial values:

```{r, fig.width = 8, fig.height = 10}
gridExtra::grid.arrange(grobs = res$`2`$plot_predict$grobs, nrow = 3, ncol = 2)
```

If a file with initial values is passed to `peaktrace()`, predictions
along the longitudinal flow path are made and visualized in a
plot. Each line in the plot represents a different scenario, e.g., the
uppermost solid lines for AMP, MAFR and MEFR represent the values of
`Q_max` in the initial file. Starting from these initial values,
predictions are made with the corresponding models, e.g.,  the first
value of the initial valus file is passed to the model that describes
the relationship of AMP between `S1` and `S2` to predict the value at
`S2`.

The initial value and the first fitted model are:

```{r}
initials[1, ]
res$`2`$models[1, ]
```

The resulting predicted value is then passed to the next model along the flow 
path, i.e., the model of `S2` and `S3` to predict the value at `S3`.

```{r}
res$`2`$models[6, ]
```

Finally, this predicted value is passed to the last model along this river 
section: the model between `S3` and `S4` to predict the value at `S4`.

```{r}
res$`2`$models[11, ]
```

This procedure is repeated for all metrics according to the initial
values file. Note that in this initial values file metric RATIO starts at
station `S2`. Therefore the first value of RATIO to predict is at station
`S3`.

The second list object refers to event type 4 (DC event). The nested
objects of this event type are shown below.

```{r}
res$`4`$settings
```

```{r}
head(res$`4`$real_AE)
```

```{r, fig.width = 8, fig.height = 3}
grid::grid.draw(res$`4`$plot_threshold)
```

```{r, fig.width = 8, fig.height = 15}
grid::grid.draw(res$`4`$plot_scatter)
```

```{r}
res$`4`$models
```

```{r, fig.width = 8, fig.height = 10}
gridExtra::grid.arrange(grobs = res$`4`$plot_predict$grobs, nrow = 3, ncol = 2)
```

## Extract AEs and perform routing with existing settings

If estimated settings are already available, it is possible to use the
settings directly to extract "real" AEs from the `Event` data. For
such an analysis, a path to a `relation` file must be provided, as
well as a path to the `Event` data and paths to `settings` and initial
values.

### Extract AEs

The following code shows the extraction of "real" AEs based on the
`settings` file which is included in the package after having been
generated with `estimate_AE()`.

```{r}
relation_path <- system.file("testdata", "relation.csv", package = "hydroroute")
events_path <- system.file("testdata", "Events", package = "hydroroute")
settings_path <- system.file("testdata", "Q_event_2_AMP-LAG_settings.csv",
	                     package = "hydroroute")
initials_path <- system.file("testdata", "initial_value_routing.csv",
                             package = "hydroroute")
real_AE <- extract_AE(relation_path, events_path, settings_path)
head(real_AE)
```

### Routing

With the extracted "real" AEs, the routing procedure can be performed to 
describe the translation and retention processes between neighboring hydrographs. 

Therefore, the output from `extract_AE()` (or similar, the output `$real_AE` from
`estimate_AE()`), the initial values data frame and the `relation` data frame have to be
passed to function `routing()`.

```{r}
relation <- utils::read.csv(relation_path)
initials <- utils::read.csv(initials_path)
res <- routing(real_AE, initials, relation)
```

This produces the same scatter plot as before when `peaktrace()` was called as the
events and the settings are the same. 

```{r, fig.width = 8, fig.height = 15}
grid::grid.draw(res$plot_scatter)
```

```{r}
res$models
```


```{r, fig.width = 8, fig.height = 10}
gridExtra::grid.arrange(grobs = res$plot_predict$grobs, nrow = 3, ncol = 2)
```

```{r, include = FALSE}
options(old)
```

## References

Greimel F, Zeiringer B, Höller N, Grün B, Godina R, Schmutz S
(2016). "A Method to Detect and Characterize Sub-Daily Flow
Fluctuations." Hydrological Processes, 30(13), 2063-2078. doi:
10.1002/hyp.10773

Greimel F, Grün B, Hayes DS, Höller N, Haider J, Zeiringer B,
Holzapfel P, Hauer C, Schmutz S (2022). "PeakTrace: Routing of
Hydropower Plant-Specific Hydropeaking Waves Using Multiple
Hydrographs - A Novel Approach." River Research and Applications,
accepted with minor revisions.
